---
title: "AMSA"
author: "Karishma Khanna, Denisse Fierro Arcos"
date: "2026-02-18"
output: 
  github_document:
    toc: true
    html_preview: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "figures/")
```

# Background
This R notebook outlines the packages needed to access and analyse NESP Data Products, as well as how to check their content structure. This notebook was originally written in Python by Karishma Kahnna It was "translated" into R by Denisse Fierro Arcos. It should be noted that although both notebooks cover the same topic, the content may vary between these notebooks as features may not be available in both Python and R.  
  
This notebook includes the following sections:  
- [Relevant R packages needed to access NESP Data Products](#relevant-packages)
- [How to access NESP Data Products](#accessing-nesp-datasets)
- [How aggregate and map NESP datasets using the H3 package](#data-aggregation-and-mapping-with-h3)
- [Identifying suitable sampling sites where trend analysis can be carried out](#identifying-suitable-trend-sites)

**Note:** The `H3` package was originally developed in C, but there are at least three packages that make this package accessible in R. We recommend you install the `h3-r`package, which published in [GitHub](https://github.com/crazycapivara/h3-r/) because this package will automatically install the original `H3` package written in C (unlike the other two options available as of February 2026). More information about the `h3-r` package, as well as a short tutorial can be found in their [documentation](https://crazycapivara.github.io/h3-r/articles/h3.html).

## Relevant packages

```{r packages}
# Connects to datasets remotely
library(arrow)
# Indexing system that uses hexagons
library(h3)
# Represents simple spatial features
library(sf)
# Data manipulation
library(dplyr)
library(lubridate)
library(tidyr)
library(stringr)
# Plotting
library(ggplot2)
# Interactive maps
library(leaflet)
```

## Accessing NESP datasets
NESP 5.9 datasets can be accessed via the `arrow` package for R. First, we will need to create a connection with the parent folder containing NESP data as follows:  

```{r}
conn <- s3_bucket("s3://data-uplift-public/", anonymous = TRUE, 
                  region = "ap-southeast-2")
```
  
Once a connection has been established, then we can select a specific dataset that interest us. In this case, we will be accessing a dataset about seagrasses. By using the `open_dataset` function from `arrow` we can load either the dataset on its entirety or a subset.

```{r}
url <- "s3://data-uplift-public/stored/datauplift/amsa/dataset"
ds <- open_dataset(sources = url, filesystem = conn)
ds
```

The AMSA dataset is partitioned (i.e., separated into parquet fragments), but it is really a single file. We can check the number of files included in a dataset as follows:  

```{r}
ds$files
```

We can check the file structure (i.e., schema) of the seabird dataset. This will give us the name of the columns contained in the dataset, which can help us filter the data before accessing.  

```{r}
ds$schema
```

From here, we can see there are 16 columns, but we can check how many records in total are included in the dataset as follows:  

```{r}
dim(ds)
```

This dataset has a few million records, which may be too much for some computers to handle, but `arrow` allows us to lazily load the dataset. This means that the data is only load into your memory when you request it, instead of loading all records at once. This could be especially useful if you have limited computing resources, so you can subset or summarise the data as you need before loading it to your computer to avoid overloading the computer's memory.  

## `H3` Aggregation and Mapping
The NESP 5.9 datasets are aggregated data products that cover large regions and time periods. It is useful to first aggregate the datasets so that we may understand their temporal-spatial bounds.

### `H3` Indexing
The `H3` library allows us to spatially index datasets at different resolutions (zoom levels) according to the record latitude and longitude.

### Visualization Strategy
- **Log-Scaling:** Maritime traffic density varies wildly (Power Law). We apply a transformation to make both ports and open oceans visible.
- **Global Max:** We calculate the maximum density across all years to create a fixed color scale. This ensures that "Red" in 2013 represents the same density as "Red" in 2024, allowing for valid comparisons.

```{r}
# --- H3 Aggregation ---
# Let's check the years available in the dataset
yrs <- ds |> 
  mutate(year = year(eventDate)) |>
  distinct(year) |>
  collect()

#Check results
yrs 
```

## Visualising the `H3` Aggregation
With the data aggregated and the global scale determined, we can now render the spatial density map. We will complete the following steps:
- **Filtering:** Selecting the specific year from the aggregated dataset.
- **Normalization:** Mapping the Log-Score to a fixed color palette using the GLOBAL_MAX_LOG calculated previously.
- **
Rendering**: Generating the 3D Hexagon Layer.

```{r}
df_short <- ds |> 
  filter(eventDate >= as_date("2013-01-01") & eventDate <= as_date("2013-12-31")) |> 
  group_by(h3Index) |> 
  summarise(n_records = n()) |> 
  collect()

df_short <- df_short |> 
  mutate(index = h3_to_geo_boundary_sf(df_short$h3Index)) |>
  ungroup()

df_short
```

```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = df_short$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = unique(df_short$n_records), 
                                         n = 10)(df_short$n_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", df_short$n_records),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```

## Regional Analysis: Sydney Harbour
While the global map uses a coarse resolution (Resolution 8, approx 0.7km²) suitable for open ocean shipping lanes, complex waterways like Sydney Harbour require finer detail.

To analyze this region effectively, we perform a Spatial Deep Dive:  
1. **Spatial Filter:** We restrict the query to the Sydney bounding box.  
2. **Re-Indexing:** We calculate a new H3 Index at Resolution 9 (approx 0.1km²). This allows us to resolve individual ferry routes and harbour traffic that would otherwise be merged into a single block.


```{r}
map_df <- ds |> 
  filter(decimalLatitude >= -33.93 & decimalLatitude <= -33.78 &
           decimalLongitude >= 151.1 & decimalLongitude <= 151.35) |> 
  filter(eventDate >= as_date("2016-01-01") & eventDate <= as_date("2016-12-31")) |> 
  select(!h3Index) |> 
  collect()

map_df <- map_df |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) 

map_df <- map_df |> 
  mutate(h3index = geo_to_h3(map_df, 12))

h3_layers <- map_df  |>
  st_drop_geometry() |> 
  group_by(h3index) |> 
  summarise(number_records = n()) 

h3_layers <- h3_layers |> 
  mutate(index = h3_to_geo_boundary_sf(h3_layers$h3index)) |>
  ungroup()
```

```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = h3_layers$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = unique(h3_layers$number_records), 
                                         n = 10)(h3_layers$number_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", h3_layers$number_records),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```


## Longitudinal Data Health
This section characterizes the dataset's temporal stability.

Long-term maritime datasets are frequently influenced by "Structural Breaks"—step-changes in data volume caused by upgrades to surveillance infrastructure (e.g., new satellite launches). The following analysis distinguishes between actual fleet growth (increased traffic) and systematic growth (improved sensor sensitivity).

### Methodology
To quantify these trends, we aggregate the data annually and derive three core metrics:  
1. **Volume:** The total count of AIS transmission records (Data Size).  
2. **Fleet:** The count of unique vessels identified by craftID (Traffic Size).  
3. **Intensity:** The average number of records per vessel (Sensor Resolution).  
  
**Normalization Strategy:** To compare trends across different scales (thousands of ships vs. millions of records), all metrics are indexed to their 2012 baseline (2012 = 100). This allows for a direct correlation analysis of relative growth rates.

```{r}
tic()
vessels_yr <-  ds |> 
  select(eventDate, craftID) |> 
  mutate(year = year(eventDate)) |> 
  distinct(year, craftID) |> 
  group_by(year) |> 
  summarise(unique_vessels = n()) |> 
  collect()
toc()

recs_yr <-  ds |> 
  group_by(year(eventDate)) |> 
  summarise(total_records = n()) |> 
  collect()

ts_yr <- ds |> 
  distinct(eventDate) |> 
  group_by(year(eventDate)) |> 
  summarise(unique_timestamps = n()) |> 
  collect()
```






