---
title: "Seagrass R Notebook"
author: "Thomas Galindo, Denisse Fierro Arcos"
date: "2026-02-10"
output: 
  github_document:
    toc: true
    html_preview: true
    always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "NESP/figures/")
```

# Background
This R notebook outlines the packages needed to access and analyse NESP Data Products, as well as how to check their content structure. This notebook was originally written in Python by Thomas Galindo. It was "translated" into R by Denisse Fierro Arcos. It should be noted that although both notebooks cover the same topic, the content may vary between these notebooks as features may not be available in both Python and R.  
  
This notebook includes the following sections:  
- [Relevant R packages needed to access NESP Data Products](#relevant-packages)
- [How to access NESP Data Products](#accessing-nesp-datasets)
- [How aggregate and map NESP datasets using the H3 package](#data-aggregation-and-mapping-with-h3)
- [Identifying suitable sampling sites where trend analysis can be carried out](#identifying-suitable-trend-sites)


## Relevant packages

```{r packages}
# Connects to datasets remotely
library(arrow)
# Indexing system that uses hexagons
library(h3)
# Data manipulation
library(dplyr)
library(tidyr)
library(stringr)
# Plotting
library(ggplot2)
# Interactive maps
library(leaflet)
```

## Accessing NESP datasets
NESP 5.9 datasets can be accessed via the `arrow` package for R. First, we will need to create a connection with the parent folder containing NESP data as follows:  

```{r}
conn <- s3_bucket("s3://data-uplift-public/", anonymous = TRUE, 
                  region = "ap-southeast-2")
```
  
Once a connection has been established, then we can select a specific dataset that interest us. In this case, we will be accessing a dataset about seagrasses. By using the `open_dataset` function from `arrow` we can load either the dataset on its entirety or a subset.

```{r}
url <- "s3://data-uplift-public/stored/datauplift/seagrass/seagrass.parquet"
ds <- open_dataset(sources = url, filesystem = conn)
ds
```

The seagrasses dataset is partitioned (i.e., separated into parquet fragments), but it is really a single file. We can check the number of files included in a dataset as follows:  

```{r}
ds$files
```

We can also check the file structure (i.e., schema) of the seagrass dataset. This will give us the name of the columns contained in the dataset, which can help us filter the data before accessing.  

```{r}
ds$schema
```

From here, we can see there are 21 columns, but we can check how many records in total are included in the dataset as follows:  

```{r}
dim(ds)
```

This dataset has over 3 million records, which may be too much for some computers to handle, but `arrow` allows us to lazily load the dataset. This means that the data is only load into your memory when you request it, instead of loading all 3 million records at once. This could be especially useful if you have limited computing resources, so you can subset or summarise the data as you need before loading it to your computer to avoid overloading the computer's memory.  

### Lazily loading dataset
If the dataset of your interest is really large, then it is probably a better idea to subset it before loading to memory. For example, we will keep records when `startYear` is 2025.  


### Loading dataset as data frame
If the dataset you are interested in is relatively small, we recommend you simply load it into memory usin the `collect` function from `arrow`.  

```{r}
# Note that running this code may take a while, so only run it if you have enough memory available
df <- ds |> 
  collect()

# We can check the first few rows in this dataset
head(df)
```

### Lazily loading dataset
If the dataset of your interest is really large, then it is probably a better idea to subset it before loading to memory. For example, we will keep records when `startYear` is 2025.  

```{r}
ds_2025 <- ds |> 
  filter(startYear == 2025) |>
  collect()

dim(ds_2025)
```

This is a much smaller version of the seagrass dataset containing on 25 records, which is much easier for a computer to handle.  

**Remember**: Use the `collect` function once you are ready to load the dataset of your interest.  


## Observation Effort, Organism Quantity and Richness
**Observation Effort** is the number of occurrence records (e.g., sampling campaigns), while **Organism Quantity** is the number of organisms of a species recorded at each occurrence.

**Observation Effort** must be taken into account when analysing trends; fluctuations in observation effort tend to distort the reported occurrences.

Species **Richness** refers to the number of different species present in a particular area. While **Organism Quantity** tells us "how many" individuals there are, **Richness** tells us "how many types" of organisms coexist.

Typically the **Richness** of an area correlates to the number of studies carried out in a given area, but eventually plateaus and this plateau is an indication that the studied region is well sampled.

## Visualising seagrass dataset
We can create a few plots to better understand the structure of this dataset.

```{r}
# Number of unique datasets per year
df |> 
  distinct(startYear, datasetName) |> 
  group_by(startYear) |> 
  count() |> 
  ggplot(aes(startYear, n))+
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks = seq(min(df$startYear, na.rm = T)-1,
                                  max(df$startYear, na.rm = T), by = 4))+
  theme_bw()+
  labs(x = "Year", y = "Number of datasets", 
       title = "Number of datasets over time", 
       subtitle = "The number of datasets fluctuates over time")+
  theme(panel.grid.minor = element_blank(), panel.border = element_blank())
```

```{r}
# Number of records per year
df |> 
  group_by(startYear) |> 
  count() |> 
  ggplot(aes(startYear, n))+
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks = seq(min(df$startYear, na.rm = T)-1,
                                  max(df$startYear, na.rm = T), by = 4))+
  scale_y_continuous(labels = scales::label_comma(),
                     breaks = seq(0, 250000, by = 50000))+
  theme_bw()+
  labs(x = "Year", y = "Number of records", 
       title = "Number of records over time", 
       subtitle = "Observation effort changes over time and is correlated to the number of datasets")+
  theme(panel.grid.minor = element_blank(), panel.border = element_blank())
```


```{r}
# Number of observations per year
df |> 
  select(startYear, occurrenceStatus, organismQuantity) |> 
  mutate(organismQuantity = case_when((is.na(organismQuantity) & 
                                         occurrenceStatus == "absent") ~ 0,
                                      (is.na(organismQuantity) &
                                         occurrenceStatus == "present") ~ 1,
                                      T ~ organismQuantity)) |>  
  group_by(startYear) |> 
  summarise(tot_obs = sum(organismQuantity, na.rm = T)) |> 
  ggplot(aes(startYear, tot_obs))+
  geom_bar(stat = "identity")+
  scale_x_continuous(breaks = seq(min(df$startYear, na.rm = T)-1,
                                  max(df$startYear, na.rm = T), by = 4))+
  scale_y_continuous(labels = scales::label_comma())+
  theme_bw()+
  labs(x = "Year", y = "Number of organisms", 
       title = "Number of organisms over time", 
       subtitle = "Number of organisms changes over time and is correlated to the number of datasets")+
  theme(panel.grid.minor = element_blank(), panel.border = element_blank())
```


```{r}
# Number of unique new species per year
df |> 
  distinct(startYear, scientificNameID) |> 
  arrange(startYear) |> 
  distinct(scientificNameID, .keep_all = T) |>
  group_by(startYear) |> 
  count() |> 
  ungroup() |> 
  mutate(tot_sp = cumsum(n)) |> 
  ggplot(aes(startYear, tot_sp))+
  geom_density(stat = "identity", fill = "lightblue", alpha = 0.5)+
  geom_point()+
  scale_x_continuous(breaks = seq(min(df$startYear, na.rm = T)-1,
                                  max(df$startYear, na.rm = T), by = 4))+
  theme_bw()+
  labs(x = "Year", y = "Species richness", 
       title = "Number of new species over time", 
       subtitle = "Number of new species is still increasing")+
  theme(panel.grid.minor = element_blank(), panel.border = element_blank())
```


We can also explore the information in this table further, by extracting information about the "Water Type" of the sites sampled.

```{r}
df_water_type <- df |> 
  mutate(short = str_extract(australianMarineRegionsTags, 
                                 "\\|(IMCRA:Mesoscale:Water Type.*Waters)\\|",
                             group = 1),
         watertype = str_extract(str_split_i(short, "\\|", i = 1), "Type:(.*)", 
                                 group = 1)) |> 
  select(!short) |> 
  drop_na(watertype) 
```

Now we can check the number of unique datasets available over time per "Water Type".

```{r}
df_water_type |> 
  distinct(startYear, datasetName, watertype) |> 
  group_by(startYear, watertype) |> 
  count() |> 
  ggplot(aes(startYear, n))+
  geom_bar(stat = "identity", aes(fill = watertype), colour = "black")+
  facet_wrap(~watertype)+
  labs(y = "Number of datasets", x = "Year")+
  theme_bw()+
  theme(legend.position = "none")
```


```{r}
df_water_type |> 
  group_by(startYear, watertype) |> 
  count() |> 
  ggplot(aes(startYear, n))+
  geom_bar(stat = "identity", aes(fill = watertype), colour = "black")+
  facet_wrap(~watertype, scales = "free_y")+
  scale_y_continuous(labels = scales::label_comma())+
  labs(y = "Number of records", x = "Year")+
  theme_bw()+
  theme(legend.position = "none")
```


```{r}
df_water_type |> 
  mutate(organismQuantity = case_when((is.na(organismQuantity) & 
                                         occurrenceStatus == "absent") ~ 0,
                                      (is.na(organismQuantity) &
                                         occurrenceStatus == "present") ~ 1,
                                      T ~ organismQuantity)) |>  
  group_by(startYear, watertype) |> 
  summarise(tot_obs = sum(organismQuantity, na.rm = T)) |> 
  ggplot(aes(startYear, tot_obs))+
  geom_bar(stat = "identity", aes(fill = watertype), colour = "black")+
  facet_wrap(~watertype, scales = "free_y")+
  labs(y = "Number of organisms", x = "Year")+
  theme_bw()+
  theme(legend.position = "none")
```


```{r}
df_water_type |> 
  distinct(startYear, watertype, scientificNameID) |> 
  arrange(startYear) |> 
  distinct(scientificNameID, watertype, .keep_all = T) |>
  group_by(startYear, watertype) |> 
  count() |> 
  group_by(watertype) |> 
  mutate(tot_sp = cumsum(n)) |>  
  ggplot(aes(startYear, tot_sp, fill = watertype, colour = watertype))+
  geom_density(stat = "identity", alpha = 0.5)+
  geom_point()+
  facet_wrap(~watertype, scales = "free_y")+
  labs(y = "Species richness", x = "Year")+
  theme_bw()+
  theme(legend.position = "none")
```


## Data aggregation and mapping with `H3`
The NESP 5.9 datasets are aggregated data products that cover large regions and temporal periods. This aggregation is useful as it helps us understand the spatio-temporal bounds of our dataset.  
We use the `H3` package in this notebook because it is an open source package that partitions the world into hexagonal cells, allowing for faster data access.

### `H3` indexing
The `H3` package allows us to spatially index datasets at different [resolutions](https://h3geo.org/docs/core-library/restable/#average-area-in-km2) (i.e., zoom levels) relative to the recorded latitude and longitude. If you go through the dataset schema, you will notice that it has an `h3Index` column, which contains unique identifiers for each hexagon in our dataset.  

```{r}
# Getting number of records for each polygon ID (h3Index)
h3_layers <- df |> 
  group_by(h3Index) |> 
  summarise(number_records = n())

# Now we get information about datasets associated to each polygon ID
h3_layers <- df |> 
  distinct(h3Index, datasetName) |> 
  group_by(h3Index) |> 
  summarise(datasets = paste(datasetName, collapse = ", ")) |> 
  right_join(h3_layers, by = "h3Index") |> 
  mutate(index = h3_to_geo_boundary_sf(h3_layers$h3Index)) |> 
  ungroup()

# We can check the results
head(h3_layers)
```


We can now create an interactive map as shown below. Note that you can interact with the map, that is zoom in and out, and hover over each title to get information about the number of records and datasets available at each polygon.  

```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = h3_layers$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = h3_layers$number_records, 
                                         n = 10)(h3_layers$number_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", h3_layers$number_records, 
                             "\nDatasets:", h3_layers$datasets),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```


## Australian Marine Regions
Each occurrence in the seagrass dataset that we have been exploring so far has been classified using the following Australian Marine Regions (see table below). Note that the classification is not 100% accurate at the region boundaries; it is accurate to ~6km.  
  
Note that the Great Barrier Reef has been included from the World Heritage Areas map.  
  
| Marine Region Type | Map Explore URL | Metadata URL (REST/About) | GeoJSON API Endpoint |
| :--- | :--- | :--- | :--- |
| **Commonwealth Marine Regions** | [Explore](https://fed.dcceew.gov.au/datasets/commonwealth-marine-regions/explore) | [Metadata](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/Commonwealth_Marine_Regions/MapServer/0) | [GeoJSON](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/Commonwealth_Marine_Regions/MapServer/0/query?outFields=*&where=1%3D1&f=geojson) |,
| **World Heritage Areas** | [Explore](https://fed.dcceew.gov.au/datasets/erin::australia-world-heritage-areas/explore) | [Metadata](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/World_Heritage_Areas/MapServer/0) | [GeoJSON](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/World_Heritage_Areas/MapServer/0/query?outFields=*&where=1%3D1&f=geojson) |
| **Marine Parks** | [Explore](https://fed.dcceew.gov.au/datasets/2b3eb1d42b8d4319900cf4777f0a83b9_0/explore) | [Metadata](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/Australian_Marine_Parks/MapServer/0) | [GeoJSON](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/Australian_Marine_Parks/MapServer/0/query?outFields=*&where=1%3D1&f=geojson) |
| **IMCRA Provincial** | [Explore](https://fed.dcceew.gov.au/datasets/erin::integrated-marine-and-coastal-regionalisation-of-australia-imcra-v4-0-provincial-bioregions/explore) | [Metadata](https://fed.dcceew.gov.au/datasets/erin::integrated-marine-and-coastal-regionalisation-of-australia-imcra-v4-0-provincial-bioregions/about) | [GeoJSON](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/IMCRA_Provincial/MapServer/0/query?outFields=*&where=1%3D1&f=geojson) |
| **IMCRA Mesoscale** | [Explore](https://fed.dcceew.gov.au/datasets/erin::integrated-marine-and-coastal-regionalisation-of-australia-imcra-v4-0-meso-scale-bioregions/explore) | [Metadata](https://fed.dcceew.gov.au/datasets/erin::integrated-marine-and-coastal-regionalisation-of-australia-imcra-v4-0-meso-scale-bioregions/about) | [GeoJSON](https://gis.environment.gov.au/gispubmap/rest/services/ogc_services/IMCRA_Mesoscale/MapServer/0/query?outFields=*&where=1%3D1&f=geojson) |
  
Follow the Explore links to explore the Australian Marine Regions. The Commonwealth Marine Regions explore map has been embedded below.  

```{r embed-map, echo=FALSE, out.width="100%"}
# Embed the interactive map URL directly
knitr::include_url("[https://fed.dcceew.gov.au/datasets/commonwealth-marine-regions/explore?location=-33.665700%2C140.517300%2C4](https://fed.dcceew.gov.au/datasets/commonwealth-marine-regions/explore?location=-33.665700%2C140.517300%2C4)", height = "600px")
```


## Exploring Available Marine Regions Tags
Australian Marine Regions Tags have the following syntax:  
  
| TAG | TYPE | OPTIONAL SUPER REGIONS | REGION |
| :--- | :--- | :--- | :--- |
| `Commonwealth Marine Region: South-east` | `Commonwealth Marine Region` | | `South-east` |
| `Marine Park:National Park Zone:Freycinet` | `Marine Park` | `National Park Zone` | `Freycinet` |
| `IMCRA:Provincial:Water Type:Cold Temperate Waters` | `IMCRA` | `Provincial`, `Water Type` | `Cold Temperate Waters` |
  
Extracting the available Australian Marine Regions Tags requires splitting the delimited column:  
  
```{r}
australian_marine_regions_tags <- df |> 
  select(australianMarineRegionsTags) |> 
  drop_na() |> 
  separate_longer_delim(australianMarineRegionsTags, delim = "|") |> 
  group_by(australianMarineRegionsTags) |> 
  summarise(count = n()) |> 
  arrange(desc(count))

# Printing the first 10 tags
head(australian_marine_regions_tags, n = 10)
```


```{r}
# Getting number of records for each polygon ID (h3Index)
marine_reg <- df |>
  drop_na(australianMarineRegionsTags) |>
  group_by(h3Index) |> 
  summarise(number_records = n())

# Now we get information about datasets associated to each polygon ID
marine_reg <- df |>
  drop_na(australianMarineRegionsTags) |>
  distinct(h3Index, datasetName) |>
  group_by(h3Index) |>
  summarise(datasets = paste(datasetName, collapse = ", ")) |>
  right_join(marine_reg, by = "h3Index") 

# Finally we get information about the marine tags
marine_reg <- df |> 
  select(h3Index, australianMarineRegionsTags) |> 
  drop_na(australianMarineRegionsTags) |>
  distinct() |> 
  separate_longer_delim(australianMarineRegionsTags, delim = "|") |> 
  distinct() |> 
  group_by(h3Index) |>
  summarise(tags = paste(australianMarineRegionsTags, collapse = ", ")) |>
  right_join(marine_reg, by = "h3Index") |> 
  mutate(index = h3_to_geo_boundary_sf(marine_reg$h3Index)) |>
  ungroup()

# Checking results
head(marine_reg)
```


```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = marine_reg$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = marine_reg$number_records, 
                                         n = 10)(marine_reg$number_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", marine_reg$number_records, 
                             "\nDatasets: ", marine_reg$datasets,
                             "\nTags: ", marine_reg$tags),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```


## Filtering to a Particular Marine Region
We could also create a map after selecting (filtering) for a particular tag as follows:  

```{r}
tropical_waters_df <- df |> 
  select(!h3Index) |> 
  filter(str_detect(australianMarineRegionsTags,
                    "IMCRA:Provincial:Water Type:Tropical Waters")) |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) 

tropical_waters_df <- tropical_waters_df |> 
  mutate(h3index = geo_to_h3(tropical_waters_df, 7))

# Getting number of records for each polygon ID (h3Index)
h3_layers <- tropical_waters_df |>
  st_drop_geometry() |> 
  group_by(h3index) |> 
  summarise(number_records = n())

# Now we get information about datasets associated to each polygon ID
h3_layers <- tropical_waters_df |>
  st_drop_geometry() |> 
  distinct(h3index, datasetName) |>
  group_by(h3index) |>
  summarise(datasets = paste(datasetName, collapse = ", ")) |>
  right_join(h3_layers, by = "h3index") |> 
  mutate(index = h3_to_geo_boundary_sf(h3_layers$h3index)) |>
  ungroup()
```



```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = h3_layers$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = unique(h3_layers$number_records), 
                                         n = 10)(h3_layers$number_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", h3_layers$number_records, 
                             "\nDatasets: ", h3_layers$datasets),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```


## Identifying suitable trend sites
In the original map that includes all data, we can see there are some sites with a high density of records, while others have very limited observations. The high density polygons are the most ideal sites for trend analysis.

### Per Region
Regions with a high density of records are typically good candidates for trend analysis. See above for subsetting and plotting by Australian Marine Regions.

### Per Dataset
Additionally, the singular datasets that comprise the aggregated data product are typically good candidates for trend analysis.

### Mapping the Spatial Extent of a Particular Dataset
Similarly to mapping Australian Marine Regions, we may map particular datasets by applying a string constraint.

### Finding Available Datasets
We may find available datasets by aggregating by the `datasetName` column.

```{r}
data_count <- df |> 
  group_by(datasetName) |> 
  summarise(count = n()) |> 
  arrange(count)

data_count
```


```{r}
data_int <- data_count |> 
  filter(count == max(count)) |> 
  pull(datasetName)
```

```{r}
trop_water_jcu_df <- df |> 
  select(!h3Index) |> 
  filter(datasetName == data_int) |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) 

trop_water_jcu_df  <- trop_water_jcu_df  |> 
  mutate(h3index = geo_to_h3(trop_water_jcu_df, 6))

# Getting number of records for each polygon ID (h3Index)
h3_layers <- trop_water_jcu_df  |>
  st_drop_geometry() |> 
  group_by(h3index) |> 
  summarise(number_records = n())

# Now we get information about datasets associated to each polygon ID
h3_layers <- trop_water_jcu_df  |>
  st_drop_geometry() |> 
  distinct(h3index, datasetName) |>
  group_by(h3index) |>
  summarise(datasets = paste(datasetName, collapse = ", ")) |>
  right_join(h3_layers, by = "h3index") |> 
  mutate(index = h3_to_geo_boundary_sf(h3_layers$h3index)) |>
  ungroup()
```

```{r}
m <- leaflet() |>
  addTiles() |>
  # Note that we use quantiles so it is easier to distinguish polygon with 
  # a high number of observations
  addPolygons(data = h3_layers$index,
              fillColor = ~colorQuantile("plasma", 
                                         domain = unique(h3_layers$number_records), 
                                         n = 10)(h3_layers$number_records),
              fillOpacity = 0.7, color = "grey", weight = 1,
              label = paste0("Records: ", h3_layers$number_records, 
                             "\nDatasets: ", h3_layers$datasets),
              labelOptions = labelOptions(style = list("max-width" = "750px",
                                                       "white-space" = "normal")))

# Plotting map
m
```




