---
title: "Plot TEMP gridded product"
author: 'E. Klein'
date: "2019-01-20"
output: 
  html_document:
    toc:  TRUE
    toc_float: TRUE
    theme: united
    highlight: tango
    code_folding: show
editor_options: 
  chunk_output_type: console
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
require(knitr)
options(width = 100, stringAsFactors=F)
opts_chunk$set(echo =T, message = F, error = F, warning = F, comment = NA,  
               fig.align = 'left',  fig.width = 7.5, fig.height = 6,
               tidy = F, cache.path = '.cache/', fig.path = 'fig/')
               
library(RColorBrewer)
palette(brewer.pal(8, "Set2"))
library(kableExtra)
library(ncdf4)
library(ggplot2)
library(dplyr)



```

Last run `r lubridate::now()`


## Introduction

This script will open a TEMP gridded timeseries product for AODN THREDDS server and plot the time series at target depths

Note that the TEMP gridded product is a rectangular array with coordinates TIME and DEPTH, being TIME the hourly timestamps and DEPTH the target depths specific to any mooring site

## Read data

First we will read the data, extract the coordinates (`TIME`, `DEPTH`) and flatten the `TEMP` array into a 1D vector. this will easy the plot of the timeseries.

It is recommended that you download the file to a local folder and then execute the code using the file. Downloading large amounts of data form the THREDDS server may cause some problems.

```{r readdata}

## this is from the AODN THREDDS server. Better to download the file to a local folder. 
fileURL = "http://thredds.aodn.org.au/thredds/dodsC/IMOS/ANMN/NSW/PH100/gridded_timeseries/IMOS_ANMN-NSW_TZ_20091029_PH100_FV02_TEMP-gridded-timeseries_END-20190828_C-20200108.nc"

nc = nc_open(fileURL)

DEPTH = ncvar_get(nc, 'DEPTH')
TIME = ncvar_get(nc, 'TIME')
TEMP = ncvar_get(nc, 'TEMP')

site_code = ncatt_get(nc, 0, "site_code")$value

```


Note that this particular file for `r site_code` site was produced interpolating temperature observations into `r length(DEPTH)` target depths:

```{r depths}
print(DEPTH)
```

We're interested in the `TEMP` variable. Note that this is a matrix of `r dim(TEMP)[1]` rows (corresponding to the target depths) and `r dim(TEMP)[2]` columns, correspondig to the hourly timestamps. In total the temperature matrix contains `r length(TEMP)` values. 

## Flatten the TEMP matrix

What we want is a data frame with `TEMP` aligned with its correpsonding `TIME` and `DEPTH`. To faltten the matrix we will use teh `as.vector` function, but this function works column-wise, so we need to transpose the matrix before flatten it.

```{r flattenTEMP}

## let's recycle the variables
TEMP_long = as.vector(t(TEMP))
print(length(TEMP_long))
```

Now, let's make the `DEPTH` and `TIME` vector that match the `TEMP` vector we've just created:

For TIME, we need to repeat the time vector as many target depths (`r length(DEPTH)`) are in the file. Also we will convert the the resulting object as a POSIXct (date-time) object:

```{r makeTIME}
TIME_long = rep(TIME, length(DEPTH))
head(TIME_long)
# time is number of day after 1950-01-01. We need it in seconds
TIME_long = as.POSIXct(TIME_long*24*60*60, origin='1950-01-01')
head(TIME_long)
```

For depth, we need to loop over each depth and repeat the target depth to match the number of timestamps:

```{r makeDEPTH}

DEPTH_long = as.numeric()
nTIMES = length(TIME)
for (depth_index in 1:length(DEPTH)){
  DEPTH_long = c(DEPTH_long, rep(DEPTH[depth_index], nTIMES))
}

```

As we now have all the vectors aligned, let's build the data frame:

```{r makeDataframe}
df = data.frame(TIME=TIME_long, DEPTH=DEPTH_long, TEMP = TEMP_long)
```

## Calculate some statistics

Let's explore the timeseries at each target depth

```{r stats}

TEMP_stats = df %>% group_by(DEPTH) %>% 
  summarise(Temp.validN = sum(!is.na(TEMP)),
            Temp.min = min(TEMP, na.rm=T),
            Temp.q05 = quantile(TEMP, 0.05, na.rm=-T), 
            Temp.q10 = quantile(TEMP, 0.05, na.rm=-T), 
            Temp.q25 = quantile(TEMP, 0.05, na.rm=-T), 
            Temp.q50 = quantile(TEMP, 0.05, na.rm=-T), 
            Temp.q75 = quantile(TEMP, 0.05, na.rm=-T), 
            Temp.q90 = quantile(TEMP, 0.05, na.rm=-T),
            Temp.q95 = quantile(TEMP, 0.05, na.rm=-T),
            Temp.max = max(TEMP, na.rm=-T), 
            Temp.mad = mad(TEMP, na.rm=T), 
            Temp.mean = mean(TEMP, na.rm=T), 
            Temp.sd = sd(TEMP, na.rm=T))

kable(TEMP_stats) %>%  kable_styling(bootstrap_options = c("striped", "hover"))
```


## Make the plot

We can now make the plot using `ggplot2` library

### Data availability coloured by temperature

```{r dataAvail, fig.height=8, fig.width=12}
pp = ggplot(df, aes(TIME, DEPTH))
pp + geom_point(size=0.2, aes(colour=TEMP)) + 
  scale_color_viridis_c() + 
  labs(subtitle = site_code) + 
  scale_y_reverse() + 
  theme_light()
```

### Time series at target depths

```{r plotTEMP, fig.height=8, fig.width=12}

pp = ggplot(df, aes(TIME, TEMP))
pp + geom_point(size=0.1) + 
  labs(subtitle = site_code) + 
  facet_grid(DEPTH~.) + 
  theme_light()

```

